Type Classes for Mathematics in Type Theory

### 2010-12-08 papers/mscs/mscs.tex
Couple more referee-guided improvements in MSCS paper.

To show that the fully unbundled approach with structures represented by predicates can be made feasible using type classes, we will tackle one by one the problems traditionally associated with their use, starting with those encountered during theorem \emph{application}. Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows\footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate. The record form is simply more convenient in that it immediately gives us named projections for laws and substructures.}:
\begin{lstlisting}
  Record SemiGroup (G: Type) (e: relation G) (op: G → G → G): Prop :=
    { sg_setoid: Equivalence e
    ; sg_ass: Associative op
    ; sg_proper: Proper (e ==> e ==> e) op }.
\end{lstlisting}
Then by (1) making \lstinline|SemiGroup| a \emph{class} (by replacing the \lstinline|Record| keyword with the \lstinline|Class| keyword), (2) marking its proofs as \emph{instances} (again by replacing a single keyword), and (3) marking the \lstinline|SemiGroup| parameter of semigroup theorems as implicit (by using curly instead of round brackets), we no longer have to pass \lstinline|SemiGroup| proofs around manually ourselves, letting instance resolution do it for us instead. Because instance resolution is part of the unifier, this also works when the statement of the theorem we wish to apply only mentions some of the components (which admittedly doesn't make much sense for semigroups).

@@

For ease of presentation we show only a proof of concept for a very concrete language. We are currently working to integrate this technique with our existing universal algebra infrastructure, in particular its data types for algebraic terms.%  Preliminary results in this area (such as a simple monoid normalization tactic) suggest yada.

>>>>

To show that the fully unbundled approach with structures represented by predicates can be made feasible using type classes, we will tackle one by one the problems traditionally associated with their use, starting with those encountered during theorem \emph{application}. Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows\footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate. The record form is simply more convenient in that it immediately gives us named projections for laws and substructures.}:
\begin{lstlisting}
  Record SemiGroup (G: Type) (e: relation G) (op: G → G → G): Prop :=
    { sg_setoid: Equivalence e
    ; sg_ass: Associative op
    ; sg_proper: Proper (e ==> e ==> e) op }.
\end{lstlisting}
Then by (1) making \lstinline|SemiGroup| a \emph{class} (by replacing the \lstinline|Record| keyword with the \lstinline|Class| keyword), (2) marking its proofs as \emph{instances} (by replacing the \lstinline|Lemma| keyword with the \lstinline|Instance| keyword), and (3) marking the \lstinline|SemiGroup| parameter of semigroup theorems as implicit (by using curly instead of round brackets), we no longer have to pass \lstinline|SemiGroup| proofs around manually ourselves, letting instance resolution do it for us instead. Because instance resolution is part of the unifier, this also works when the statement of the theorem we wish to apply only mentions some of the components (which admittedly doesn't make much sense for semigroups).

@@

For ease of presentation we show only a proof of concept for a very concrete language. We are currently working to integrate this technique with our existing universal algebra infrastructure. In particular, the latter's term data type should be ideally suited to serve as a generic symbolic representation of terms in a wide class of algebras. This should let us implement the basic setup of the technique once and for all, so that quotation for new algebraic structures can be enabled with minimal effort.
  %  Preliminary results in this area (such as a simple monoid normalization tactic) suggest yada.
